---
title: "Untitled"
author: "Teshager"
date: "December 10, 2020"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(rmarkdown)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```



# Activating libraries

```{r}
library(MASS)
library(stats)
library(base)
library(Rcpp)
library(usethis)
library(dplyr)
library(ggplot2)
library(gstat)
library(maps)
library(sp)
library(geos)
library(rgdal)
library(PROJ)
library(sf)
library("CARBayesST")
library(Matrix)
library(readxl)
library(parallel)
library(dplyr)
library(spData)
library(spdep)
library(foreach)
library(rgdal)
library(GGally)
library("INLA")
```


# Impotring data into R

```{r}
library(DT)
Dat_week<- read_xlsx("C:/Users/Amin/Desktop/Data_T/Teshager weekly final data 2.xlsx")
Dat_week<-Dat_week %>% mutate(MPR=(Dat_week$TM/Dat_week$Population)*10000, Tcos=cos(3.14*Dat_week$T/6),Tsin=sin(3.14*Dat_week$T/6))
Dat_week$Woreda_Nam<-Dat_week$Woreda
Dat_week$Lands_temp<-Dat_week$Land_Surface_temp-273.15
head(Dat_week)

```





# Importing the shapefiles of Amhara regions

```{r}
library(sf)
s<-read_sf("C:/Users/Amin/Desktop/Data_T/Administration/Amhara_woreda_recent_boundary.shp")
sh_po <- st_read("C:/Users/Amin/Desktop/Data_T/Administration/Amhara_woreda_recent_boundary.shp")
st_geometry_type(sh_po)
st_crs(sh_po)
st_bbox(sh_po)
ggplot() + 
  geom_sf(data = sh_po, size =1, color = "black", fill = "gray") + 
  ggtitle("Amhara region Boundary map between 1990-2020") + 
  coord_sf()
```


# Changing coordinate systems


```{r}
library(dplyr)
library(MASS)
library(raster)
s.sf.gcs <- st_transform(sh_po, "+proj=longlat +datum=WGS84")
st_crs(s.sf.gcs)
crs(s.sf.gcs)
```


# Take care of considering the Lake Tana data in analysis which may distort the results of the analysis.

### Removing or deleting data which satisfies a certain conditions 



# ploting data

```{r}
 ggpairs(Dat_week, columns=c(9, 30:41))
```




```{r}
attach(Dat_week)
ggplot(Dat_week,aes(x=factor(`Budget Year`),y=TM))+geom_boxplot(fill="blue",alpha=0.8)+
  scale_x_discrete(name="Physical year")+
  scale_y_continuous(name="Malaria cases per week")+
  theme(text=element_text(size = 16),plot.title=element_text(size=18,face="bold"))

```


## Grouping by woreda names

```{r}
library(sf)
group_IG <- group_by(Dat_week, Woreda_Nam)
MPR.av <- summarise(group_IG, TM.mean = mean(TM))
s.sf.gcs$Tmal<-MPR.av$TM.mean
s.sf.gcs$Tmal
```


# Mapping the spatial data
 # GGHB.IG <- spTransform(GGHB.IG, CRS("+proj=longlat +datum=WGS84 +no_defs"))
 #Then a map of SMR can be drawn using the following code.

```{r}
 library(rgdal)

  library(leaflet)
  colours <- colorNumeric(palette = "YlOrRd", domain = s.sf.gcs$Tmal)
  map1 <- leaflet(data=s.sf.gcs) %>%
  addTiles() %>%
  addPolygons(fillColor = ~colours(Tmal), weight=1, color="",
      fillOpacity = 0.7) %>%
    addLegend(pal = colours, values = s.sf.gcs$Tmal, opacity = 2,
     title="MPR") %>%
   addScaleBar(position="bottomleft")
 map1
```



## Generating the spatial weight matrix
```{r}
 library("spdep")
 W.nb <- poly2nb(s.sf.gcs, row.names = MPR.av$Woreda_Nam)
 W.list <- nb2listw(W.nb, style = "B")
 W <- nb2mat(W.nb, style = "B")
 
```



# spatiotemporal model

```{r}
 form <- TM ~ Elevation +Land_Surface_temp + NDVI + Spec_humudity +Max_tempertaure + Min_temperature +Rain_fall + Soil_moisture +offset(log(Population))
                     
model1 <- glm(formula = form, family = "poisson",data = Dat_week,offset = log(Population))
 resid.glm <- residuals(model1)
 summary(model1)$coefficients
 summary(model1)$dispersion
print(model1)
coef(model1)
logLik(model1)

 

```




# Modeling with out covariate






```{r}
length(Elevation)
length(Land_Surface_temp)
length(Soil_moisture)
length(Rain_fall)
length(Min_temperature)
length(Max_tempertaure)
length(NDVI)
length(Spec_humudity)
length(TM)
length(Population)

```



```{r}
library("CARBayesST")
set.seed(123456)
formula2<- TM ~ Elevation+Land_Surface_temp+Soil_moisture+Rain_fall+Min_temperature+Max_tempertaure+Spec_humudity+NDVI+offset(log(Population))
chain3<- ST.CARanova(formula=formula2, family = "poisson",
data = Dat_week, W=W, burnin =180000, n.sample = 300000, thin =100)

```



```{r}
print(chain3)
```
```{r}
logLik(chain3)
```


```{r}
summary(chain3)
```

```{r}
summary(chain3$samples)
```
# fitted values

```{r}
Yfitted<-fitted.values(chain3)
write.csv(Yfitted,"C:/Users/Amin/Desktop/Data_T/fitted_value2.csv",row.names = FALSE)
```

# Residuals 
```{r}
res<-residuals(chain3)
write.csv(res,"C:/Users/Amin/Desktop/Data_T/residuals.csv",row.names = FALSE)
```

# Model fit

```{r}
chain3$modelfit
```

```{r}
A<-chain3$samples
```

# Trace Plot of parameters

```{r}
library(mcmc)
library(coda)
par(mfrow=c(5,2))
b.sample<-mcmc.list(chain3$samples$beta)
K<-b.sample[,1:9]
plot(K,col="red")
```


# Accept: acceptance probability for the parameters

```{r}
chain3$accept
```

```{r}
chain3$samples
```


