---
title: "Paper2"
author: "Teshager"
date: "3/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Paper II: Project 

## Project 2

```{r}
M_Dat<- read.csv("D:/Thesis final Data/Final edited data final/Teshager monthly malaria final data 2.csv")
M_Dat$Woreda_Nam<-M_Dat$Woreda
M_Dat$SMR<-(M_Dat$TM/M_Dat$Population)*10000
M_Dat$T_cos<-cos((3.14*M_Dat$Time)/6)
M_Dat$T_sin<-sin((3.14*M_Dat$Time)/6)
M_Dat$Month_1<-M_Dat$Time
M_Dat$Month_2<-M_Dat$Time
M_Dat$Month_3<-M_Dat$Time
M_Dat$FiD1<-M_Dat$FID+1
M_Dat$FiD2<-M_Dat$FID+1
M_Dat$Pop<-log(M_Dat$Population)
M_Dat$FiD.T<-seq(1:14592)
head(M_Dat)
```


## Activating the library
```{r}
library(foreach)
library(usethis)
library(spData)
library(PROJ)
library(geos)
library(sf)
library(sp)
library(spdep)
library(Matrix)
library(parallel)
library(tibble)
library(inlatools)
library(dplyr)
library(leaflet)
library(ggplot2)
library(INLA)
library(RgoogleMaps)
library(bitops)
library(rjson)
library(ggmap)
library(ggsn)
library(classInt)
library(viridis)
library(dplyr)
library(RColorBrewer)
library(reshape2)
library(tidyr)
library(dplyr)
library(remotes)
library(ggregplot)
```



### Importing the Shapefile
```{r}
library(spData)
library(sp)
library(PROJ)
library(geos)
library(sf)
library(spdep)
shapef <- st_read("D:/Administration/Amhara_woreda_recent_boundary.shp")
```


### Generating spatial weight matrix in R-InLA 
```{r}
library(spData) 
library(spdep)
temp <- poly2nb(shapef)
nb2INLA("Amhara.graph", temp)
Amhara_shape <- paste(getwd(),"/Amhara.graph",sep="")
Amhara_shape
```


## Main Effects
```{r}
Type_I_Main<-TM~T_sin+T_cos+Elevatn+Max_temp+Min_temp+Landsurf_temp+Soil_mois+Rain_fall+NDVI+Relative.humidity 
Type_I_main<-inla(formula = Type_I_Main, family ="nbinomial",data=M_Dat,offset=Pop,     control.predictor=list(compute=TRUE),control.compute=list(dic=TRUE,waic=TRUE))
```


## Type I interaction: Model specification 

```{r}
modI<-TM~T_sin+T_cos+Elevatn+Min_temp+Max_temp+Landsurf_temp+Rain_fall+Soil_mois+NDVI+Relative.humidity+LLIN_ATLEASTONE+
  f(FiD1,model="bym",graph=Amhara_shape,hyper = list(prec.unstruct=list(prior="loggamma",param=c(1,0.001)),prec.spatial=list(prior="loggamma",param=c(1,0.001)))) +
f(Month_1,model="rw2") +
f(Month_2,model="iid") +
f(FiD.T,model="iid")

```


# General poisson regression Models
```{r}
ZNB_I<-inla(formula =modI, family ="gpoisson",data=M_Dat,offset=Pop,     control.predictor=list(compute=TRUE),control.compute=list(config=TRUE,cpo=TRUE,dic=TRUE,waic=TRUE))
```


## The AR1 models for the temporal models
```{r}
modI_AR<-TM~T_sin+T_cos+Elevatn+Min_temp+Max_temp+Landsurf_temp+Rain_fall+Soil_mois+NDVI+Relative.humidity+LLIN_ATLEASTONE+f(FiD1,model="bym",graph=Amhara_shape,hyper = list(prec.unstruct=list(prior="loggamma",param=c(1,0.001)),prec.spatial=list(prior="loggamma",param=c(1,0.001)))) +
f(Month_1,model="ar1") +
f(Month_2,model="iid") +
f(FiD.T,model="iid")
GPOi_I_AR<-inla(formula =modI_AR, family ="gpoisson",data=M_Dat,offset=Pop,     control.predictor=list(compute=TRUE),control.compute=list(config=TRUE,cpo=TRUE,dic=TRUE,waic=TRUE))
```


## default prior

## The AR1 models for the temporal models
```{r}
modI_rw1<-TM~T_sin+T_cos+Elevatn+Min_temp+Max_temp+Landsurf_temp+Rain_fall+Soil_mois+NDVI+Relative.humidity+LLIN_ATLEASTONE+f(FiD1,model="bym",graph=Amhara_shape) +
f(Month_1,model="rw1") +
f(Month_2,model="iid") +
f(FiD.T,model="iid")
GPOi_I_rw1<-inla(formula =modI_rw1, family ="gpoisson",data=M_Dat,offset=Pop,     control.predictor=list(compute=TRUE),control.compute=list(config=TRUE,cpo=TRUE,dic=TRUE,waic=TRUE))
```




# Zoroiflated poisson regression Models
```{r}
ZNB_I<-inla(formula =modI, family ="zeroinflatedpoisson0",data=M_Dat,offset=Pop,     control.predictor=list(compute=TRUE),control.compute=list(config=TRUE,cpo=TRUE,dic=TRUE,waic=TRUE))
```

# Zoroiflated Negative Binomial regression Models
```{r}
ZINB_I_ZINB<-inla(formula =modI, family ="zeroinflatednbinomial0",data=M_Dat,offset=Pop,     control.predictor=list(compute=TRUE),control.compute=list(config=TRUE,cpo=TRUE,dic=TRUE,waic=TRUE))
```


```{r}

plot(ZNB_I, plot.fixed.effects=FALSE, plot.lincomb=FALSE, plot.random.effects=FALSE,
     plot.hyperparameters=FALSE, plot.predictor=FALSE, plot.q=FALSE, plot.cpo=TRUE,
     single=FALSE)
```


## The final model selected for analysis is generalized poisson model and the analysis is presented 
using different statistics.

```{r}
summary(ZNB_I)
```



### The spatial random effects and excessive risks
```{r}
GP<-ZNB_I$marginals.random[[1]]
zeta.GP1 <- unlist(lapply(GP,function(x)inla.emarginal(exp,x)))
#zeta.ST1
Area_spec<-zeta.GP1[1:152]
Spa_struc<-zeta.GP1[153:304]
ID<-seq(1:152)
Woreda_Nam<-M_Dat$woreda[1:152]
Teshe_2<-data.frame(ID,Woreda_Nam,Area_spec,Spa_struc)
Teshe_2$FID<-Teshe_2$ID
Teshe_2<-as.data.frame(Teshe_2)
head(Teshe_2)
write.csv(Teshe_2,file="C:/Users/DTU/Desktop/spatial 2.csv")
Tem_ID<-seq(1:96)
GP2<- ZNB_I$marginals.random[[2]]
TS <- unlist(lapply(GP2,function(x)inla.emarginal(exp,x)))
GP3<- ZNB_I$marginals.random[[3]]
TU <- unlist(lapply(GP3,function(x)inla.emarginal(exp,x)))
Tempo<-data.frame(Tem_ID,TS,TU)
Tempo<-as.data.frame(Tempo)
write.csv(Tempo,file="C:/Users/DTU/Desktop/Tempo.csv")
Tempo
```



```{r}
temporal.rw2 <- lapply(ZNB_I$marginals.random$Month_1,
function(X){marg <- inla.tmarginal(function(x) exp(x), X)
inla.emarginal(mean, marg)})

temporal.IID <- lapply(ZNB_I$marginals.random$Month_2,
function(X){
marg <- inla.tmarginal(function(x) exp(x), X)
inla.emarginal(mean, marg)
})

```


#Exporting nonlinear trends of time (structural and unstructral temporal component)
```{r}
T1<-as.matrix(temporal.rw2)
T2<-as.matrix(temporal.IID)
T<-cbind(T1,T2)
colnames(T) <- c("Temporal_rw2", "Temporal_IID")
rownames(T)<-NULL
T
write.csv(T,file="C:/Users/DTU/Desktop/temp.csv")
```


Min_temp+Max_temp+Landsurf_temp+Rain_fall+Soil_mois+NDVI+Relative.humidity
# Computing lagged values of valariables
```{r}
M<-M_Dat
attach(M)
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(MinT1=dplyr::lag(Min_temp,n=1,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(MinT2=dplyr::lag(Min_temp,n=2,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(MaxT1=dplyr::lag(Max_temp,n=1,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(MaxT2=dplyr::lag(Max_temp,n=2,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(LST1=dplyr::lag(Landsurf_temp,n=1,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(LST2=dplyr::lag(Landsurf_temp,n=2,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(RF1=dplyr::lag(Rain_fall,n=1,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(RF2=dplyr::lag(Rain_fall,n=2,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(SM1=dplyr::lag(Soil_mois,n=1,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(SM2=dplyr::lag(Soil_mois,n=2,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(NDVI1=dplyr::lag(NDVI,n=1,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(NDVI2=dplyr::lag(NDVI,n=2,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(RH1=dplyr::lag(Relative.humidity,n=1,default = NA))%>%as.data.frame()
M<-M %>%
  group_by(woreda)%>%dplyr::mutate(RH2=dplyr::lag(Relative.humidity,n=2,default = NA))%>%as.data.frame()
```


```{r}
head(M)
write.csv(M,"C:/Users/DTU/Downloads/ccc.csv")
```
```{r}
a<-glm(TM~Elevatn+Min_temp+MinT1+MinT2+Max_temp+MaxT1+MaxT2+NDVI1+NDVI2+Relative.humidity+RH1+RH2+Soil_mois+SM1+SM2+Rain_fall+RF1+RF2+Landsurf_temp+LST1+LST2+LLIN_ATLEASTONE,data=M,offset=Pop, family = quasipoisson(link="log"))
summary(a)
```


```{r}
m.s<-step(a,trace=0)
```



### Exessive Risk and 

```{r}
a <- 0
prob.csi <- lapply(GP, function(x) {1 - inla.pmarginal(a, x)})
prob.ex.risk<-prob.csi[1:152]
Ex.risk<-as.matrix(prob.ex.risk)
#colnames(Ex.risk)<-"Excessive_risk"
#rownames(Ex.risk)<-NULL
#Ex.risk2<-as.data.frame(Ex.risk)

```

```{r}
Ex.risk<-as.data.frame(Ex.risk)
Ex.risk$Woreda_Nam<-M_Dat$woreda[1:152]
write.csv(Ex.risk,"C:/Users/DTU/Desktop/UncerSR.csv")
```

### Generating the fixed values

```{r}
fit<-ZNB_I$summary.fitted.values
fit$Woreda<-M_Dat$Word_Nm
fit$wor<-M_Dat$woreda
fit$period<-M_Dat$Mn_GC_1
fit$FiD<-M_Dat$FID
fit$Time<-M_Dat$Time
head(fit)
write.csv(fit,"C:/Users/DTU/Desktop/Fitted.csv")
```

```{r}
intract<-ZNB_I$marginals.random[[4]]
zeta.ST1 <- unlist(lapply(intract,function(x)inla.emarginal(exp,x)))
a <- 0
PP <- lapply(intract, function(x) {1 - inla.pmarginal(a, x)})
PP<-as.matrix(PP)
#colnames(Ex.risk)<-"Excessive_risk"
#rownames(Ex.risk)<-NULL
#Ex.risk2<-as.data.frame(Ex.risk)
PP
```

```{r}
PP$Mon<-M_Dat$Mn_GC_1
PP$year<-M_Dat$year
write.csv(PP,"C:/Users/DTU/Desktop/PP.csv")
```

```{r}
intract<-ZNB_I$marginals.random[[4]]
Int <- unlist(lapply(intract,function(x)inla.emarginal(exp,x)))
Int<-as.data.frame(Int)
head(Int)
write.csv(Int,"C:/Users/DTU/Desktop/interac.csv")
```






