---
title: "Eyu 29 2014"
author: "Teshager"
date: "5/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



### Libraries
```{r}
library(Rcpp)
library(ggplot2)
library(GGally)
library(bmstdr)
library(foreach)
library(usethis)
library(Matrix)
library(parallel)
library(sp)
library(spdep)
library(sf)
library(spData)
#library(spDataLarge)
library(INLA)
library(CARBayes)
library(CARBayesST)
library(shapefiles)
library(ggplot2)
library(truncdist)
library(rstan)
library(dplyr)
library(tidyverse)
library(rgdal)
library(dplyr)
library(inlabru)
library(MASS)
```



### Import data
```{r}
Dat <- read.csv("C:\\Users\\DTU\\Desktop\\Machine Learning data and code 2\\Monthly malaria cases ST sorted NA estimated.csv")
Dat$FID<-Dat$FID +1
Dat$logEcases<- log(Dat$Population)
#Dat$T_cos<-cos((3.14*Dat$Month_new)/6)
#Dat$T_sin<-sin((3.14*Dat$Month_new)/6)
Dat$T_cos<-cos((3.14*Dat$Time)/6)
Dat$T_sin<-sin((3.14*Dat$Time)/6)
M_shap<-read_sf("D:/Administration/Amhara_woreda_recent_boundary.shp")
M_shap$FID<-M_shap$FID1
Dat$Malrate<-(Dat$TM/Dat$Population)*1000
Dat$Mal_disc<-ceiling(Dat$Malrate) 
Dat$Mal_incid <- Dat$TM/Dat$Population
Dat$mon_year<-as.Date(Dat$mon_yr,"%m/%d/%Y")
head(Dat)
```


## Data preProcessing ... standadizing variables

```{r}
DD<-Dat
DD<-DD %>% mutate_at(c('Elevatn', 'Soil_mois','Landsurf_temp','Rain_fall','Min_temp', 'Max_temp', 'NDVI','Relative.humidity','LLIN_ATLEASTONE','Soil_mois_1','Soil_mois_2','Landsurf_temp_1','Landsurf_temp_2','Rain_fall_1','Rain_fall_2','Min_temp_1','Min_temp_2','Max_temp_1', 'Max_temp_2','NDVI_1','NDVI_2','Relative.humidity_1', 'Relative.humidity_2'),~(scale(.) %>% as.vector))
Dat_2<-DD
head(Dat_2)
```




```{r}
ptime <- ggplot(data=Dat_2,  aes(x=factor(Time), y=Malrate)) +
  geom_boxplot() +
  labs(x = "Month", y = "Number of cases per 100,000")  +
  stat_summary(fun=median, geom="line", aes(group=1, col="red")) +
  theme(legend.position = "none")
ptime
```





# Spatio-temporal models 

```{r}
F2 <- Malrate ~ T_sin+T_cos+Elevatn+Max_temp+Min_temp+Rain_fall+Landsurf_temp+Relative.humidity+Soil_mois+NDVI+LLIN_ATLEASTONE

```

```{r}
F22 <- Malrate ~ year + T_sin+T_cos+ Elevatn + Max_temp + Min_temp + Rain_fall + Landsurf_temp + Relative.humidity + Soil_mois + NDVI + LLIN_ATLEASTONE

Mod_ST_4 <- Malrate~ year+Month_new+ Max_temp + Min_temp + Soil_mois + NDVI + LLIN_ATLEASTONE + Rain_fall + Landsurf_temp + Elevatn + Relative.humidity + Soil_mois_1+ Landsurf_temp_1 + Min_temp_1 + Max_temp_1+Relative.humidity_1+NDVI_1+Rain_fall_1

Mod_ST_41 <- Malrate~ T_cos+T_sin+ Max_temp + Min_temp + Soil_mois + NDVI + LLIN_ATLEASTONE + Rain_fall + Landsurf_temp + Elevatn + Relative.humidity+ Soil_mois_1 + Landsurf_temp_1 + Min_temp_1 + Max_temp_1 + Relative.humidity_1+NDVI_1+Rain_fall_1+Rain_fall_2+ Min_temp_2 + Max_temp_2 + NDVI_2 + Relative.humidity_2 + Soil_mois_2 + Landsurf_temp_2
```


### Data partition 

```{r}
set.seed(1234)
tn <- 96
sn <- 152
#blcok cross validation 
valids<- c(1:30)
#valids <- sample(1:sn,size=16)
#validt <- sort(sample(1:tn, size=30))
vrows <- getvalidrows(sn=sn, tn=tn, valids=valids, allt=TRUE)

valids2<- c(1:152)
#validt <- sort(sample(78:tn, size=19))
validt <- sort(sample(69:tn, size=28))
vrs <- getvalidrows(sn=sn, tn=tn, valids=valids2,validt = validt)
```


#### Modeling spatiotemporal data 
# a. Point refernced data
# 1. Spatio-temporal model fitting with the spTimer package

##ST prediction using spTimer

### M1: Linear 

```{r}
M1<-Bsptime(model = "lm",formula = Mod_ST, data=Dat_2,scale.transform = "SQRT", validrows = vrows, N=5000)
summary(M1)
M2<-Bsptime(model = "lm",formula = Mod_ST, data=Dat_2,scale.transform = "NONE", validrows = vrows, N=5000)
summary(M2)
```
```{r}
M1_1<-Bsptime(model = "lm",formula =F22, data=Dat_t,scale.transform = "SQRT", validrows = vrs, N=5000)
summary(M1_1)
```

```{r}
M11<-Bsptime(model = "lm",formula = F2, data=Dat_t,scale.transform = "SQRT", validrows = vrows, N=5000)
summary(M11)
M11_1<-Bsptime(model = "lm",formula =F22, data=Dat_t,scale.transform = "SQRT", validrows = vrows, N=5000)
summary(M11_1)
```





```{r}
#M1<-Bsptime(model = "lm",formula = F2, data=Dat,scale.transform = "SQRT", validrows = vrows, N=5000)
#summary(M1)
M1_1_22<-Bsptime(model = "lm",formula =Mod_ST_41, data=Dat,scale.transform = "SQRT", validrows = vrows_1, N=5000)
summary(M1_1_22)
```


## M2: Separable model
```{r}
#M2<-Bsptime(model="separable", formula=F2, data=Dat,scale.transform = "SQRT",coordtype = "lonlat",coords = 26:27, validrows = vrows, N=5000 )
#summary(M2)

M2_1<-Bsptime(model="separable", formula=Mod_ST, data=Dat,scale.transform = "NONE",coordtype = "lonlat",coords = 26:27, validrows = vrows, N=5000 )
summary(M2_1)
```




### M3: Independent GP model with nugget efffect: Temporally independent GP for the spatiotemporal process w(si,t)
# using a spTimer package with a Matern correlation function 

```{r}
DFit<-Dat_2[1:11704,]
Dfore<-Dat_2[11705:14592,]
library(spTimer)
M3_1 <- Bsptime(package="spTimer", formula=F2, data=DFit,coordtype="utm", coords=26:27, scale.transform = "SQRT", N=5000)
#M3_2 <- Bsptime(package="spTimer", formula=F22, data=DFit,coordtype="utm", coords=26:27, scale.transform = "SQRT", N=5000)
nfore <- predict(M3_1$fit,newdata=Dfore, newcoords= ~Lngtd_X + Lattd_Y, type="temporal", foreStep=19, tol=0.05)
```

### M4: Marginal model implementaion using stan for independent GP model


```{r}
library(rstan)
M4_1<- Bsptime(package = "stan",model = "GP", formula=Mod_ST_4, data=Dat, scale.transform = "NONE",
            coordtype = "lonlat",coords = 26:27, validrows=vrows_1, N=10,burn.in =2,mchoice=T,verbose=F)
summary(M4_1)
```





## M5: Autoregressive model (hierarchical AR models usng spTimer package, RA(1))


```{r}
#M5_1<-Bsptime(package = "spTimer",model = "AR", formula=F2, data=Dat, scale.transform = "NONE",
 #           coordtype = "lonlat",coords = 26:27, validrows=vrows, N=5000 )
M5<-Bsptime(package = "spTimer",model = "AR", formula=Mod_ST_4, data=Dat, scale.transform = "SQRT",
            coordtype = "lonlat",coords = 26:27, validrows=vrows, N=5000 )
summary(M5)
```





##M6. AR model using InLa

```{r}

```




### M7. Spatiotemporal dynamic models

```{r}
library(spTDyn)
F3 <- TM ~ Elevatn+Max_temp+tp(Min_temp)+Rain_fall+sp(Rain_fall)+Landsurf_temp+Relative.humidity+sp(Relative.humidity)+Soil_mois+NDVI+LLIN_ATLEASTONE
M7<-Bsptime(package = "sptDyn",model ="GP",formula = F3, data=Dat, coordtype = "lonlat",coords = 26:27,scale.transform = "NONE",validrows = vrows,N=5000)
summary(M7)
```


### M8. All paraMETERS ARE SPATIALLY VARYING , 7.5.2 dynamic ST model and regression coefficients allowed to have a dynamic prior distributions


```{r}
M8<-Bsptime(package = "spBayes",formula=F2, data=Dat, scale.transform = "NONE",prior.sigma2 = c(2,25),prior.tau2 = c(2,25),prior.sigma.eta = c(2,0.01),mchoice = TRUE,
            coordtype = "lonlat",coords = 26:27, validrows=vrows, N=5000 )
summary(M8)
```


### M9. ST models based on Gaussian predictive process (GPP)

```{r}
M9 <- Bsptime(package = "spTimer", model="GPP", g_size=12, formula = F2, data=Dat,coordtype = "lonlat",coords = 26:27,validrows =vrows, scale.transform = "NONE")
summary(M9)

#M9_1 <- Bsptime(package = "spTimer", model="GPP", g_size=5, formula = F2, data=D_U500,coordtype = "lonlat",coords = 26:27,validrows =vrows, scale.transform = "NONE")
#summary(M9_1)
```





# Spatiotemporal Areal data 

### Creating spatial contiguity matrix



```{r}
W_queen <- poly2nb(pl=M_shap, queen = TRUE)
W_Rook <- poly2nb(pl=M_shap, queen = FALSE)
WEIT=nb2listw(neighbours=W_Rook,style="W")
WM <- nb2mat(neighbours=WEIT$neighbours,style="B", zero.policy = TRUE)
```


```{r}
library(spData)
library(spdep)
temp<-poly2nb(M_shap)
nb2INLA("Amhara.graph",temp)
Amhara_shape<-paste(getwd(),"/Amhara.graph",sep = "")
Amhara_shape
```




# Spatio-temporal models 

```{r}
F_disc <- Mal_disc ~ T_sin+T_cos+year+Elevatn+Max_temp+Min_temp+Rain_fall+Landsurf_temp+Relative.humidity+Soil_mois+NDVI+LLIN_ATLEASTONE
```

## Modeling statistic areal level data  # poisson model
```{r}
ptime <- ggplot(data=Dat,  aes(x=factor(Time), y=Malrate)) +
  geom_boxplot() +
  labs(x = "Month", y = "Number of cases per 100,000")  +
  stat_summary(fun=median, geom="line", aes(group=1, col="red")) +
  theme(legend.position = "none")
ptime
```







### Running the spatiotemporal data 

```{r}
Mod_ST <- Mal_disc~ T_sin +T_cos+ Max_temp + Min_temp + Soil_mois + NDVI + LLIN_ATLEASTONE + Rain_fall + Landsurf_temp + Elevatn +Relative.humidity 
```


```{r}
Mod_ST_1 <- Malrate~ year+Month_new+ Max_temp + Min_temp + Soil_mois + NDVI + LLIN_ATLEASTONE + Rain_fall + Landsurf_temp + Elevatn +Relative.humidity+Soil_mois_1+Landsurf_temp_1+Min_temp_1+Max_temp_1+Relative.humidity_1+NDVI_1+Rain_fall_1
Mod_ST_2 <- Malrate~ T_cos+T_sin+ Max_temp + Min_temp + Soil_mois + NDVI + LLIN_ATLEASTONE + Rain_fall + Landsurf_temp + Elevatn +Relative.humidity+Soil_mois_1+Landsurf_temp_1+Min_temp_1+Max_temp_1+Relative.humidity_1+NDVI_1+Rain_fall_1
```


```{r}
set.seed(12345)
vs <- sample(nrow(Dat), 0.2*nrow(Dat))
```








## 1. ST_linear model

```{r}
ST_linear <- Bcartime(formula=F2, data=Dat_2, scol="FID", tcol= "Time",  
                      W=WM, model="linear", family="gaussian", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10, 
                       validrows=vrs, verbose=F)
summary(ST_linear)
```


```{r}
ST_linear <- Bcartime(formula=F_disc, data=Dat_2, scol="FID", tcol= "Time",  
              W=WM, model="linear", family="poisson", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10, validrows = vrs,
                        verbose=F)
summary(ST_linear)
```

```{r}
ST_linear_poi<- Bcartime(formula=F_disc, data=Dat_2, scol="FID", tcol= "Time",  
                       W=WM, model="linear", family="poisson", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10, validrows=vrs,
                       verbose=F,link = "log")
summary(ST_linear_poi)
```





## 2.ST_Autoregressive models 
  
### 2.1 AR=1 model for temporal effects

```{r}
set.seed(12345)

M2_AR1_1 <- Bcartime(formula=Mod_ST, data=Dat_2, scol="FID", tcol= "Time",  
                       W=WM, model="ar", AR=1, family="poisson", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10, rseed = 12345, validrows=vrs)
summary(M2_AR1_1)
```
,
                      
### 2.2 ST_AR2

```{r}
set.seed(5)

ST_AR2_val <- Bcartime(formula=Mod_ST, data=Dat, scol="FID", tcol= "Time",  
                       W=WM, model="ar", AR=2, family="poisson", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10, rseed = 12345,validrows=vrs)
summary(ST_AR2_val)
```


### 2.3 ST_Poisson_AR1
```{r}
ST_PAR1 <- Bcartime(formula=Mod_ST, data=Dat, scol="FID", tcol= "Time",  
                       W=WM, model="ar", AR=1, family="poisson", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10, 
                       validrows=vrs, verbose=F,rseed = 12345)
summary(ST_PAR1)
```

### 2.3.1 ST_ZIP_AR1
```{r}
ST_AR_ZIP <- Bcartime(formula=Mod_ST, data=Dat, scol="FID", tcol= "Time",  
                       W=WM, model="ar", AR=1, family="zip", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10, 
                       validrows=vrs, verbose=F,rseed = 12345)
summary(ST_AR_ZIP)

```

### 2.4 ST_Poisson_AR2
```{r}
ST_PAR2 <- Bcartime(formula=Mod_ST, data=Dat_2, scol="FID", tcol= "Time",  
                       W=WM, model="ar",AR=2, family="poisson", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10, rseed=12345,
                       validrows=vrs)
summary(ST_PAR2)
```





# 3. ST_ANOVA gaussian

```{r}
ST_ANOVA <- Bcartime(formula=F2, data=Dat_2, scol="FID", tcol= "Time",  
                       W=WM, model="anova", family="gaussian", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10, interaction = FALSE,
                       validrows=vrs, verbose=F)
summary(ST_ANOVA)
```


### 3.1 ST_ANOVA -poisson without interaction
 
```{r}
ST_anova_2<- Bcartime(formula=Mod_ST, data=Dat, scol="FID", tcol= "Time",  
                       W=WM, model="anova", family="poisson", package="CARBayesST", validrows=vrs,
                       N=100000, burn.in=80000, thin=10, interaction = FALSE,link = "log")
summary(ST_anova_2)
```


### 3.3 ST_ANOVA with interaction gaussian
```{r}
ST_ANOVA_33 <- Bcartime(formula=F2, data=Dat_2, scol="FID", tcol= "Time",  
                       W=WM, model="anova", family="gaussian", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10, interaction = TRUE,
                       validrows=vrs, verbose=F)
summary(ST_ANOVA_33)
```


### 3.4 ST_ANOVA -poisson with interaction of time and space
 
```{r}
ST_anova_34<- Bcartime(formula=Mod_ST, data=Dat, scol="FID", tcol= "Time",  
                       W=WM, model="anova", family="poisson", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10,validrows=vrs, interaction = TRUE,link = "log")
summary(ST_anova_34)
```



# 4. ST_sepspatial

### 4.1 ST_separable model-gaussian

```{r}
ST_sep <- Bcartime(formula=F2, data=Dat_2, scol="FID", tcol= "Time",  
                       W=WM, model="sepspatial", family="gaussian", package="CARBayesST", 
                      validrows=vrows, N=100000, burn.in=80000, thin=10)
summary(ST_sep)
```


### 4.2 ST_sepspatial_poisson

```{r}
ST_sep_2 <- Bcartime(formula=Mod_ST, data=DFit, scol="FID", tcol= "Time",  
                       W=WM, model="sepspatial", family="poisson", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10,link = "log")
summary(ST_sep_2)
```










## INLA ##

```{r}
m_1<-c("bym","rw1")
m_2<-c("bym","iid")
```


### II. INLA poisson

```{r}
M_INLA_val <- Bcartime( formula=Mod_ST, data=Dat_2,adj.graph = Amhara_shape, scol ="FID", tcol="Time",model=c("bym","rw2"), family="zeroinflatedpoisson0", package="inla", N=1000, burn.in=400, thin = 10,link = "log",validrows = vrows) 
```
```{r}
summary(M_INLA_val)
```


### Poisson

```{r}
Poin_INLA <- Bcartime( formula=Mod_ST, data=Dat_2,adj.graph = Amhara_shape, scol ="FID", tcol="Time",model=m_1, family="poisson", package="inla", N=1000, burn.in=400, thin = 10,link = "log",validrows = vrs) 
```

```{r}
summary(Poin_INLA)
```


### Zero-inflated poisson

```{r}
model=c("bym","ar1")
ZIP_INLA<- Bcartime( formula=Mod_ST, data=Dat_2,adj.graph=Amhara_shape, scol ="FID", tcol="Time",  
                         model=model, family="zeroinflatedpoisson0", package="inla", link="log",validrows=vrs) 
```




```{r}
summary(ZIP_INLA)
```

### negative binomial


```{r}
NB_INLA<- Bcartime( formula=Mod_ST, data=Dat_2,W=WM, scol ="FID", tcol="Time",  
                        offsetcol=NULL,  model=m_1, family="nbinomial", package="inla", link="log",  N=1000, burn.in=400, thin = 10,validrows=vrs) 
```
```{r}
summary(NB_INLA)
```



```{r}
Gpoisson_INLA<- Bcartime( data=Dat_2,formula=Mod_ST,W=WM, scol ="FID", tcol="Time",  
                        offsetcol=NULL, model=c("bym","ar1"),link="log", family="gpoisson", package="inla",N=1000, burn.in=400, thin = 10,validrows=vrs) 
```


```{r}
summary(Gpoisson_INLA)
```

### ZIP using "zip" function in inla

```{r}
ZIN_INLA<- Bcartime( data=Dat_2,formula=Mod_ST,W=WM, scol ="FID", tcol="Time",  
                        offsetcol=NULL, model=c("bym","ar1"),link="log", family="zeroinflatednbinomial0", package="inla",N=1000, burn.in=400, thin = 10,validrows=vrs) 
```

```{r}
summary(ZIN_INLA)
```





########## Trying INLA using a new data set

### Importing the data

```{r}

Dat_2 <- read.csv("C:\\Users\\DTU\\Desktop\\Machine Learning data and code 2\\Dat_2.csv")

F2 <- Malrate ~ T_sin+T_cos+Elevatn+Max_temp+Min_temp+Rain_fall+Landsurf_temp+Relative.humidity+Soil_mois+NDVI+LLIN_ATLEASTONE
Mod_ST <- Mal_disc~ T_sin +T_cos+ Max_temp + Min_temp + Soil_mois + NDVI + LLIN_ATLEASTONE + Rain_fall + Landsurf_temp + Elevatn +Relative.humidity 


Mod_ST_inc <- Mal_incid~ T_sin +T_cos+ Max_temp + Min_temp + Soil_mois + NDVI + LLIN_ATLEASTONE + Rain_fall + Landsurf_temp + Elevatn +Relative.humidity 

tn=96
sn=152
valids=c(1:152)
validt <- sort(sample(78:tn, size=19))
vs_1<- getvalidrows(sn=sn,tn=tn,valids=valids, validt=validt)
dfit <- Dat_2[-vs_1, ]
dfore <- Dat_2[vs_1, ]


DFit<-Dat_2[1:11704,]
Dfore<-Dat_2[11705:14592,]
```



```{r}
library(spTimer)
M3_1 <- Bsptime(package="spTimer", formula=F2, data=DFit,coordtype="utm", coords=26:27, scale.transform = "SQRT", N=5000)
#M3_2 <- Bsptime(package="spTimer", formula=F22, data=DFit,coordtype="utm", coords=26:27, scale.transform = "SQRT", N=5000)
nfore <- predict(M3_1$fit,newdata=Dfore, newcoords= ~Lngtd_X + Lattd_Y, type="temporal", foreStep=19, tol=0.05)
```



```{r}
M_INLA_II<- Bcartime( formula=Mod_ST, data=Dat_2,W=WM, scol ="FID", tcol="Time",  
                        offsetcol=NULL,  model=c("bym","ar1"), family="nbinomial", package="inla", link="log",validrows=vrs,  N=1000, burn.in=400, thin = 10,rseed = 12345) 

```



```{r}
ST_sep_22 <- Bcartime(formula=Mod_ST_inc, data=Dat_2, scol="FID", tcol= "Time",  
                       W=WM, model="ar",AR=1, family="gaussian", package="CARBayesST", 
                       N=100000, burn.in=80000, thin=10,
                        verbose=F)
```

Dat_2$L_expect2<-Dat_2$Expected/1000
Dat_2$L_expect3<-ceiling(Dat_2$L_expect2)
